Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-26 projectReport/abstract.tex
Found too many small errors

\begin{abstract}
    It is a well-established practice to collect measurements from underlying artificial, natural, and physical systems through observables in order to realise a dynamical model with greater descriptive power. 
    Many approaches to forecasting exist but they necessitate an understanding of systems we often don't know much about.

>>>>

\begin{abstract}
    It is a well-established practice to collect measurements from underlying artificial, natural, and physical systems through observables to realise a dynamical model with greater descriptive power. 
    Many approaches to forecasting exist but they necessitate an understanding of systems we often don't know much about.

### 2022-01-26 projectReport/chapter1.tex
Rereading the project. On 5.4
All G edits done
Found too many small errors

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. 
Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. 
For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. 
However, our world is much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually `negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. We encounter many difficultiesin the process of modelling such systems:

@@

Models derived primarily from data can be classified into three categories: 
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item  Interpretable models (i.e., they establish relationships between internal physical mechanisms), 
  \item Partially interpretable models capturing some modes of the dynamics, 
  \item Non-interpretable models, defined as such mainly due to the fact that they are defined on a different phase space that is usually high-dimensional. 
\end{enumerate}

@@

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding method of delay-coordinates~\cite{takens1981detecting} when one has `good' observations from the system; this learnt system could then be used to forecast the observed data. 
Takens delay embedding theorem~\cite{takens1981detecting} and its various generalisations (see,~\cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the understanding of the dynamics unreliable (see \cite{principe1992prediction}). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter~\ref{ch3}), it can cause predictions to fail.

>>>>

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. 
Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. 
For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. 
However, our world is much more complex than that which can be distilled into elegant equations. 
We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually `negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. 
We encounter many difficulties in the process of modelling such systems:

@@

Models derived primarily from data can be classified into three categories.
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item  Interpretable models (i.e., they establish relationships between internal physical mechanisms), 
  \item Partially interpretable models capturing some modes of the dynamics, 
  \item Non-interpretable models, defined as such mainly because they are defined on a different phase space that is usually high-dimensional. 
\end{enumerate}

@@

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding method of delay-coordinates~\cite{takens1981detecting} when one has `good' observations from the system; this learnt system could then be used to forecast the observed data. 
Takens delay embedding theorem~\cite{takens1981detecting} and its various generalisations (see,~\cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the understanding of the dynamics unreliable (see~\cite{principe1992prediction}). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter~\ref{ch3}), it can cause predictions to fail.

### 2022-01-26 projectReport/chapter2.tex
Found too many small errors

At its most elementary level, a dynamical system is just something that evolves deterministically through time. In the context of this project, deterministic refers to the fact that a system evolves according to specified rules rather than based on random events. 
Dynamical systems arise in a variety of situations. A continuous-time dynamical system describes the states for all values of the time. Specifically, if the motion of a pendulum in which the quantities such as the angular position and angular momentum are known at all times, then it is a continuous-time dynamical system. The equations of the dynamical system can take the form of one or more ordinary differential equations that determine the relevant quantities at any future time if we know the initial location and momentum. 
In ecology, discrete-time dynamical systems are widely used to model population growth. The model in this case is a function calculating the following generation's population given the population of the previous generation. If we know the starting population, we may once again calculate the population at any time in the future.

@@

A core concept in the study of dynamical systems is that of invariance. 
\begin{Definition}
  [\bf Invariant Set]\label{Dfn_InvariantSet}\rm
  Given a discrete-time dynamical system $T: U \to U$, a non-empty subset $A \subset U$ is said to be an \emph{invariant set} if $T(A) =A$. 
\end{Definition}
We also define the orbit of a function $T$.
\begin{Definition}
  [\bf Orbit of $T$]\label{Dfn_Orbit}\rm
  The orbit of $T$ is defined  is to the sequence $\bar{u} = {\{u_n\}}_{n\in \mathbb{Z}}$ obeying the update equation, $u_{n+1}=Tu_n$, $n \in \mathbb{Z}$. 
\end{Definition}

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^{pu}\}$, where $T^{p+1}u=u$ for some $p\in\mathbb{Z}$.  The entire space $U$ could be also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$.

@@

\begin{Example}\label{ex_halfu} \rm
If $U=[0,1]$ has map $Tu= u/2$, then the only invariant set is $\{0\}$, and all orbits approach this invariant set as time flows in the forward direction. 
Indeed, if a some non-zero invariant set (call it $B$) exists, then there is  some $r\in(0,1]\cap{B}$. But $r\notin{T(B)}$ since any orbit with initial value $r$ will be a decreasing sequence. Moreover, every orbit of T will be decreasing and therefore approach the value $0$ as $n\rightarrow\infty$.
\end{Example}

One may ask if every orbit approaches an invariant set? In general the answer is no, since for the dynamical system $T: \mathbb{R} \to \mathbb{R}$ defined by $Tu=2u$, any orbit that does not intersect the invariant set $\{0\}$, will not approach any invariant set.

@@

\begin{Example}\label{ex_Tu2}\rm
  For the map $Tu=u^2$ defined on $[0,1]$ all orbits lie in the invariant set $\{1\}$ or else would would approach the invariant set $\{0\}$. 
\end{Example}

Invariant sets have various properties. Invariant sets can be attracting or repelling depending on how orbits in their vicinity behave. 
Recall the examples $Tu =u/2$ and $Tu=2u$ defined on $\mathbb{R}$ (where $\{0\}$ `attracts' orbits) or the example, $Tu=u^2$ defined on $[0,1]$ (where $\{1\}$ `repels' orbits). 
We are interested in attractive invariant sets since they capture the long-term dynamics as time increases. 
In particular, we are interested in those invariant sets named attractors.

@@

Indeed in our previous example~\ref{ex_halfu, ex_Tu2}, it can be verified that the system  $Tu=u/2$ defined on  $\mathbb{R}$ and $Tu=u^2$ defined on $[0,1]$  the singleton set $\{0\}$ is an attractor.  For the system,  $Tu=1-|2u-1|$ on $[0,1]$, one may easily verify that the only attractor is the entire space $[0,1]$. This follows from the fact that between any two points $u$ and $v$ s.t. $u< v$ in $[0,1]$, and for any $a,b$ so that $u\le a < b \le v$, we can find an $n$ so that $T^n(a,b)=[0,1]$  as is explained in Ex.~\ref{ex_standardTentMap}.

@@

The dynamics restricted to an invariant set can be complicated. For instance an invariant set could be just a single point or it could have an infinite set. If the invariant set is infinite, then complicated dynamics are possible. A particular, well-studied phenomenon of such complexity gives rise to so-called chaotic behaviour, a subject studied in detail over the past fifty years.

@@

\begin{Definition}
  [\bf {(Sensitive Dependence on Initial Conditions)}]\label{Dfn_SDIC}\rm
A dynamical system $T: U \to U$ is said to have \textit{sensitive dependence on initial conditions (SDIC)} if there exists a $\delta > 0$ such that for every $u \in U$ and in every neighborhood of $u \in U$ there exists a $v\in{U}$ and an integer $N:=N{(u,v)}\in\mathbb{Z}$ such that $d(T^Nu,T^Nv)>\delta$. 	
\end{Definition}

@@

To this end, we make 3 remarks:
\vspace{-5mm}
\begin{enumerate}
  \item The $\delta>0$ in the definition of SDIC is independent of $u$. 
  \item In every neighborhood of $u$, we may not necessarily find all points $v$ in the neighborhood distinct from $u$ that would separate from the forward iterates of $u$. 
  \item  $N$ depends upon $u$ and $v$ chosen, and their iterates may not separate forever (i.e., for all $n>N$) and we allow their iterates to get arbitrarily close in the future. 
\end{enumerate}

@@

\begin{Example}\label{ex_LM} \rm
  The logistic map (LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$, is a classic example used to illustrate the chaotic behaviour that a system may exhibit. When $r=4$, the one-dimensional recurrence relation $x_{n+1}=4x_n(1-x_n)$ can be used as the kindergarten's model to exhibit the presence of SDIC.
  In the plot Fig.~\ref{fig:log_sdic} below, we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$. 
  The values of $x_n$ are plotted agains time $n$  where they have initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.

@@

\begin{Definition}
  [\bf {Topologically Transitive}]\label{Dfn_TopolTrans}\rm
	A dynamical system $F: U \to U$ is topologically transitive if for any pair of nonempty open sets $E_1$ and $E_2$ there exists a $n\in\mathbb{N}$ such that $$T^n(E_1) \cap E_2 \not= \emptyset$$. 
\end{Definition}

@@

\begin{Definition}
  [\bf {Meager Set}]\label{Dfn_Meager Set}\rm
A subset of a topological space $U$ is said to be a meager set if it can be written as a countable union of sets of with empty interior. The complement of the a meager set is said to be a residual set.
\end{Definition}

@@

  \begin{proof}
    The graph of the map $T$ is piecewise linear with two straight lines, one connecting the points $(0,0)$ and $(\frac{1}{2},1)$ and the other connecting $(\frac{1}{2},1)$ with $(1,0)$. They form a so-called tent with base centered at $u=1/2$. The graph of the map $T^2$ comprises two symmetric tents with their base centered at $1/4$ and $3/4$. See Figure~\ref{fig:T2tentmap}

@@

    In general, the graph of  $T^k$ contains $2^{k-1}$ tents.
    Given any point $u$ and an open neighborhood $W \subset[0,1]$ of $u$, we can find $k\in\mathbb{Z}$ large enough so that $T^k(W) = [0,1]$ because we can accommodate a tent whose base is contained in $W$.
    This implies the existence of two points $w_1$ and $w_2$ in $W$ so that $d(T^kw_1,T^kw_2)= 1$, and by the triangle inequality  at least one of the inequalities $d(T^ku,T^kw_1)> \delta$ or $d(T^ku,T^kw_2)> \delta$ hold when $\delta\in (0,\frac{1}{2})$.  So $T$ has sensitive dependence on initial conditions.

@@

When $S$ is conjugate to $T$, the dynamics of the two systems are in some way `dynamically equivalent'. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (see,~\cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

>>>>

At its most elementary level, a dynamical system is just something that evolves deterministically through time. In the context of this project, deterministic refers to the fact that a system evolves according to specified rules rather than based on random events. 
Dynamical systems arise in a variety of situations. A continuous-time dynamical system describes the states for all values of the time. Specifically, if the motion of a pendulum in which the quantities such as the angular position and angular momentum are always known, then it is a continuous-time dynamical system. The equations of the dynamical system can take the form of one or more ordinary differential equations that determine the relevant quantities at any future time if we know the initial location and momentum. 
In ecology, discrete-time dynamical systems are widely used to model population growth. The model in this case is a function calculating the following generation's population given the population of the previous generation. If we know the starting population, we may once again calculate the population at any time in the future.

@@

A core concept in the study of dynamical systems is that of invariance. 
\begin{Definition}
  [\bf Invariant Set]\label{Dfn_InvariantSet}\rm
  Given a discrete-time dynamical system $T: U \to U$, a non-empty subset $A \subset U$ is said to be an \emph{invariant set} if $T(A) =A$. 
\end{Definition}
We also define the orbit of a function $T$.
\begin{Definition}
  [\bf Orbit of $T$]\label{Dfn_Orbit}\rm
  The orbit of $T$ is defined to be the sequence $\bar{u} = {\{u_n\}}_{n\in \mathbb{Z}}$ obeying the update equation, $u_{n+1}=Tu_n$, $n \in \mathbb{Z}$. 
\end{Definition}

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^{pu}\}$, where $T^{p+1}u=u$ for some $p\in\mathbb{Z}$.  The entire space $U$ could also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$.

@@

\begin{Example}\label{ex_halfu} \rm
If $U=[0,1]$ has map $Tu= u/2$, then the only invariant set is $\{0\}$, and all orbits approach this invariant set as time flows in the forward direction. 
Indeed, if some non-zero invariant set (call it $B$) exists, then there is  some $r\in(0,1]\cap{B}$. But $r\notin{T(B)}$ since any orbit with initial value $r$ will be a decreasing sequence. Moreover, every orbit of T will be decreasing and therefore approach the value $0$ as $n\rightarrow\infty$.
\end{Example}

One may ask if every orbit approaches an invariant set. In general the answer is no, since for the dynamical system $T: \mathbb{R} \to \mathbb{R}$ defined by $Tu=2u$, any orbit that does not intersect the invariant set $\{0\}$, will not approach any invariant set.

@@

\begin{Example}\label{ex_Tu2}\rm
  For the map $Tu=u^2$ defined on $[0,1]$ all orbits lie in the invariant set $\{1\}$ or else would approach the invariant set $\{0\}$. 
\end{Example}

Invariant sets have various properties. Invariant sets can be attracting or repelling depending on how orbits in their vicinity behave. 
Recall the examples $Tu =u/2$ and $Tu=2u$ defined on $\mathbb{R}$ (where $\{0\}$ `attracts' orbits) or the example, $Tu=u^2$ defined on $[0,1]$ (where $\{1\}$ `repels' orbits). 
We are interested in attractive invariant sets since they capture the long-term dynamics as time increases. 
We are interested in those invariant sets named attractors.

@@

Indeed, in our previous example~\ref{ex_halfu},~\ref{ex_Tu2}, it can be verified that the system  $Tu=u/2$ defined on  $\mathbb{R}$ and $Tu=u^2$ defined on $[0,1]$  the singleton set $\{0\}$ is an attractor.  For the system,  $Tu=1-|2u-1|$ on $[0,1]$, one may easily verify that the only attractor is the entire space $[0,1]$. This follows from the fact that between any two points $u$ and $v$ s.t. $u< v$ in $[0,1]$, and for any $a,b$ so that $u\le a < b \le v$, we can find an $n$ so that $T^n(a,b)=[0,1]$  as is explained in Ex.~\ref{ex_standardTentMap}.

@@

The dynamics restricted to an invariant set can be complicated. For instance, an invariant set could be just a single point or it could have an infinite set. If the invariant set is infinite, then complicated dynamics are possible. A well-studied phenomenon of such complexity gives rise to so-called chaotic behaviour, a subject studied in detail over the past fifty years.

@@

\begin{Definition}
  [\bf {(Sensitive Dependence on Initial Conditions)}]\label{Dfn_SDIC}\rm
A dynamical system $T: U \to U$ is said to have \textit{sensitive dependence on initial conditions (SDIC)} if there exists a $\delta > 0$ such that for every $u \in U$ and in every neighbourhood of $u \in U$ there exists a $v\in{U}$ and an integer $N:=N{(u,v)}\in\mathbb{Z}$ such that $d(T^Nu,T^Nv)>\delta$. 	
\end{Definition}

@@

To this end, we make 3 remarks:
\vspace{-5mm}
\begin{enumerate}
  \item The $\delta>0$ in the definition of SDIC is independent of $u$. 
  \item In every neighbourhood of $u$, we may not necessarily find all points $v$ in the neighbourhood distinct from $u$ that would separate from the forward iterates of $u$. 
  \item  $N$ depends upon $u$ and $v$ chosen, and their iterates may not separate forever (i.e., for all $n>N$) and we allow their iterates to get arbitrarily close in the future. 
\end{enumerate}

@@

\begin{Example}\label{ex_LM} \rm
  The logistic map (LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$, is a classic example used to illustrate the chaotic behaviour that a system may exhibit. When $r=4$, the one-dimensional recurrence relation $x_{n+1}=4x_n(1-x_n)$ can be used as the kindergarten's model to exhibit the presence of SDIC.
  In the plot Fig.~\ref{fig:log_sdic} below, we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$. 
  The values of $x_n$ are plotted against time $n$  where they have initial values $x_0=0.2$ (blue) and $x_0=0.20001$ (red). Initially the two trajectories overlap, but they diverge completely at $n=12$ whereafter they follow distinct paths.

@@

\begin{Definition}
  [\bf {Topologically Transitive}]\label{Dfn_TopolTrans}\rm
	A dynamical system $F: U \to U$ is topologically transitive if for any pair of nonempty open sets $E_1$ and $E_2$ there exists a $n\in\mathbb{N}$ such that \[T^n(E_1) \cap E_2 \not= \emptyset.\]
\end{Definition}

@@

\begin{Definition}
  [\bf {Meagre Set}]\label{Dfn_Meagre Set}\rm
A subset of a topological space $U$ is said to be a meagre set if it can be written as a countable union of sets of with empty interior. The complement of the meagre set is said to be a residual set.
\end{Definition}

@@

  \begin{proof}
    The graph of the map $T$ is piecewise linear with two straight lines, one connecting the points $(0,0)$ and $(\frac{1}{2},1)$ and the other connecting $(\frac{1}{2},1)$ with $(1,0)$. They form a so-called tent with base centred at $u=1/2$. The graph of the map $T^2$ comprises two symmetric tents with their base centred at $1/4$ and $3/4$. See Figure~\ref{fig:T2tentmap}

@@

    In general, the graph of  $T^k$ contains $2^{k-1}$ tents.
    Given any point $u$ and an open neighbourhood $W \subset[0,1]$ of $u$, we can find $k\in\mathbb{Z}$ large enough so that $T^k(W) = [0,1]$ because we can accommodate a tent whose base is contained in $W$.
    This implies the existence of two points $w_1$ and $w_2$ in $W$ so that $d(T^kw_1,T^kw_2)= 1$, and by the triangle inequality  at least one of the inequalities $d(T^ku,T^kw_1)> \delta$ or $d(T^ku,T^kw_2)> \delta$ hold when $\delta\in (0,\frac{1}{2})$.  So $T$ has sensitive dependence on initial conditions.

@@

When $S$ is conjugate to $T$, the dynamics of the two systems are in some way `dynamically equivalent'. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (see,~\cite{de2013elements}) is a more extensive system capturing all the essential dynamics of its factor.

### 2022-01-26 projectReport/chapter3.tex
Found too many small errors

Takens embedding theorem does not take into account the noise, and to describe it we construct a multidimensional observable using the method of stacking previous observations.
The concept of an observable is understood in the sense of an observable as originally introduced by~\cite{takens1981detecting, genericObservableAeyels} to refer to a probe or measurement function inserted into the system. 
When speaking of stacking, we mean that a delay-coordinate map is defined by
$\Phi_{k,\theta}(w) := (\theta(T^{-k}w)\ldots,\theta(T^{-1}w),\theta(w))$.  
In essence, Takens theorem refers to the fact that for a sufficiently large $k$, we can define a dynamical system on the space $\mathbb{R}^{k+1}$ with states evolving as $\Phi_{k,\theta}(w)$, $\Phi_{k,\theta}(Tw)$, $\Phi_{k,\theta}(T^2w)$, etc. This dynamical system is topologically conjugate to the unknown underlying system $(W,T)$. We recall the Takens delay embedding theorem next.

@@

Takens Theorem states a result establishing a relationship between the observed and underlying dynamical systems by showing that the concatenation of a sufficiently large number of previous observations into a vector will, under certain conditions, generate a map between the vectors from the respective systems.  We formulate the theorem from \cite{takens1981detecting} after establishing preliminary definitions.

@@

Takens Theorem states a result establishing a relationship between the observed and underlying dynamical systems by showing that the concatenation of a sufficiently large number of previous observations into a vector will, under certain conditions, generate a map between the vectors from the respective systems.  We formulate the theorem from~\cite{takens1981detecting}.

@@

% The Takens embedding theorem does not guarantee global dissipativity.
% \ednote{B:Links with~\ref{subs_LearnGamma} in discussion of dissipativity of $\Gamma$.  Sir, I'm not sure how to write this here - I think it a beneficial point to make so as to set the stage for the later discussion in ~\ref{subs_LearnGamma}, but I am unsure there as well.}

>>>>

Takens embedding theorem does not consider the noise, and to describe it we construct a multidimensional observable using the method of stacking previous observations.
The concept of an observable is understood in the sense of an observable as originally introduced by~\cite{takens1981detecting, genericObservableAeyels} to refer to a probe or measurement function inserted into the system. 
When speaking of stacking, we mean that a delay-coordinate map is defined by
$\Phi_{k,\theta}(w) := (\theta(T^{-k}w)\ldots,\theta(T^{-1}w),\theta(w))$.  
In essence, Takens theorem refers to the fact that for a sufficiently large $k$, we can define a dynamical system on the space $\mathbb{R}^{k+1}$ with states evolving as $\Phi_{k,\theta}(w)$, $\Phi_{k,\theta}(Tw)$, $\Phi_{k,\theta}(T^2w)$, etc. This dynamical system is topologically conjugate to the unknown underlying system $(W,T)$. We recall the Takens delay embedding theorem next.

@@

Takens Theorem states a result establishing a relationship between the observed and underlying dynamical systems by showing that the concatenation of a sufficiently large number of previous observations into a vector will, under certain conditions, generate a map between the vectors from the respective systems.  We formulate the theorem from~\cite{takens1981detecting} after establishing preliminary definitions.

@@

@@

% The Takens embedding theorem does not guarantee global dissipativity.
